# Presenting the Data {#ANOVAsummaries}

When a research question involves the relationship between two or more variables, such as comparing the mean response across levels of a factor, successful presentations of the data which address the question of interest _partition the variability_.  This key idea is essential to both the data presentation and the data analysis.

We have already argued that variability makes addressing questions difficult.  If every subject had the same response to a particular exposure, there would be no need for statistics.  We would simply evaluate one subject and determine which treatment to give.  Statistics exists because of the ambiguity created by variability in the responses.  In response to this variability, our statistical graphics and models distinguish (partition) the various sources of variability.  That is, with any analysis, we try to answer the question "why aren't all the values the same?  What are the reason(s) for the difference we are observing?"

From the [Organic Food Case Study](#CaseOrganic), consider the primary question of interest:

  > Is there evidence of a relationship between the type of food a person is exposed to and their moral expectations, on average, following exposure?
  
What we are really asking is "does the food exposure help explain the differences in the moral expectations of individuals?"  We know that there are differences in moral expectations between individuals.  But, are these differences solely due to natural variability (some people are just inherently, possibly due to how they were raised, more or less liberal with moral beliefs); or, is there some systematic component that explains at least a portion of the differences between individuals.  We are thinking about partitioning the "why the responses differ" (the variability).

A good graphic must then tease out how much of the differences in the moral expectations is from subject-to-subject variability and how much is due to the food exposure.  First, consider a common graphic which is __not__ useful in this situation (Figure \@ref(fig:anovasummaries-bad-bar)).

```{r anovasummaries-organic-data, echo=FALSE, ref.label="organic-data"}
```

```{r anovasummaries-bad-bar, echo=FALSE, fig.cap="Illustration of a poor graphic using the Organic Food Case Study; the graphic does not give us a sense of variability.  As a result, it is not clear how different these means really are."}
ggplot(data = organic.df,
       mapping = aes(y = moral_avg, x = Food_Condition)) +
  stat_summary(fun.y = "mean", geom = "bar") +
  stat_summary(aes(label=round(..y..,2)), fun.y = "mean", geom = "label") +
  labs(x = "Food Exposure Group",
       y = "Average Moral Expectation Score") +
  theme_bw(12)
```

To determine an appropriate graphic, we need to remember that we want to partition the variability. So, we must not only compare the differences between the groups but also allow the viewer to get a sense of the variability within the group.  A common way of doing this within the engineering and sciences is to construct side-by-side boxplots, as illustrated in Figure \@ref(fig:anovasummaries-organic-boxplot).

```{r anovasummaries-organic-boxplot, echo=FALSE, fig.cap="Comparison of the moral expectations for college students exposed to different types of food."}
ggplot(data = organic.df,
       mapping = aes(y = moral_avg, x = Food_Condition)) +
  geom_boxplot() +
  labs(x = "Food Exposure Group",
       y = "Moral Expectation Score") +
  theme_bw(12)
```

From the graphic, we see that the moral expectation scores seem to have nearly the same pattern in each of the exposure groups.  More, the center of each of the groups is roughly the same.  That is, there does not appear to be any evidence that the type of food to which a subject is exposed is associated with moral expectations, on average.  

Side-by-side boxplots can be helpful in comparing large samples as they summarize the location and spread of the data.  When the sample is smaller, it can be helpful to overlay the raw data on the graphic in addition to the summary provided by the boxplot.  We might also consider adding additional information, like the mean within each group.  An alternative to boxplots is to use violin plots which emphasize the shape of the distribution instead of summarizing it like boxplots.  Yet another option is to construct density plots which are overlayed on one another.  This works when there are only a small number of groups; if the number of groups is large, then placing the distributions side-by-side is much more effective.  A comparison of these approaches is in Figure \@ref(anovasummaries-organic-comparison).  

```{r anovasummaries-organic-comparison, echo=FALSE, fig.cap="Multiple ways to effectively compare the response across multiple groups."}
p1 <- ggplot(data = organic.df,
             mapping = aes(y = moral_avg, x = Food_Condition)) +
  geom_boxplot() +
  stat_summary(aes(group = 1), fun.y = "mean", geom = "line", size = 1.25) +
  stat_summary(fun.y = "mean", geom = "point", size = 4, shape = 17) +
  labs(x = "Food Exposure Group", y = "Moral Expectation Score", 
       title = "Boxplot with Means") +
  theme_bw(12)

p2 <- ggplot(data = organic.df,
             mapping = aes(y = moral_avg, x = Food_Condition)) +
  geom_jitter(height = 0, width = 0.2, alpha = 0.75, colour = "grey75") +
  geom_boxplot(alpha = 0.5, size = 1.1) +
  labs(x = "Food Exposure Group", y = "Moral Expectation Score",
       title = "Boxplot with Raw Data") +
  theme_bw(12)

p3 <- ggplot(data = organic.df,
             mapping = aes(y = moral_avg, x = Food_Condition)) +
  geom_violin(draw_quantiles = c(0.5)) +
  stat_summary(fun.y = "mean", geom = "point", size = 4, shape = 17) +
  labs(x = "Food Exposure Group", y = "Moral Expectation Score",
       title = "Violin Plots") +
  theme_bw(12)

p4 <- ggplot(data = organic.df,
             mapping = aes(x = moral_avg, colour = Food_Condition,
                           fill = Food_Condition)) +
  geom_density(alpha = 0.25, size = 1.1) +
  labs(x = "Moral Expectation Score", y = "Density",
       fill = "", colour = "",
       title = "Overlayed Densities") +
  theme_bw(12) + 
  theme(legend.position = "bottom")

gridExtra::grid.arrange(p1, p2, p3, p4, nrow = 2, ncol = 2)
```

Each of these plots is reasonable.  What makes them useful in addressing the question is that in each plot, we can compare the degree to which the groups differ relative to the variability within a group.  That is, we partition the variability.  With each plot, we can say that one of the reasons the groups differ is because of exposure to different food types; however, this difference is extremely small relative to the fact that regardless of which food group you were exposed to, the variability in moral expectations with that group is quite large.  Since the predominant variability in the moral exposure is the variability within the groups, we would say there is no signal here.  That is, there is no evidence that the average scores differ across food exposure groups.

The key to a good summary is understanding the question of interest and building a graphic which addresses this question through a useful characterization of the variability.
