# Summarizing Multivariable Data {#Regsummaries}

For the [Seismic Activity Case Study](#CaseGreece), we are primarily interested in characterizing the relationship between bracketed duration and the magnitude of the earthquake.  As we discussed in the previous chapters, this broad question could be refined into a question falling into one of three categories:

  - Marginal Relationship: overall, do two variables tend to move together (are they correlated)?
  - Isolation of Effect: does a relationship exist after accounting for the effect of additional variables?  Or, what is the effect "above and beyond" the effect of additional variables?
  - Interplay: how does the relationship between two variables change as a result of a third?

As always, the key is developing summaries which help to address the question of interest.  
  

## Characterizing the Marginal Relationship of Two Quantitative Variables
Suppose we are interested in the following question:

  > In general, does the bracketed duration increase as the magnitude increases?
  
This question is about the overall relationship between these two quantitative variables.  Graphically, we can examine the relationship between these two variables using a _scatter plot_.  The response is placed on the y-axis and the predictor along the x-axis.  Figure \@ref(fig:regsummaries-magnitude) illustrates the relationship between the bracketed duration and the magnitude.

```{r regsummaries-magnitude, echo=FALSE, fig.cap="Relationship between the bracketed duration and the magnitude of an earthquake for locations Greece."}
ggplot(data = greece.df,
       mapping = aes(x = Magnitude, y = BD02)) + 
  geom_point() +
  labs(x = "Moment Magnitude of the Earthquake",
       y = "Bracketed Duration (s) at Measurement Location") + 
  theme_bw(12)
```

The graphic highlights several components of the relationship.  First, we note that as the magnitude of the event increases, the bracketed duration also tends to increase.  This is intuitive --- as the size of the earthquake increases, the length of time the ground shakes with extreme force increases.  Again, this is not universally true; there are cases for which the magnitude was high, but the bracketed duration was lower.  Our goal is to characterize the overall trend.  We also notice that as the magnitude increases, the variability in the bracketed duration also tends to increase.  That is, for earthquakes of small magnitudes, it seems fairly easy to anticipate the bracketed duration; however, the bracketed duration is much more difficult to anticipate for larger magnitudes.

A nice visual tool when exploring the relationship between two quantitative variables is a _smoothing spline_.  The details of its construction are beyond the scope of this text, but we can think of it as representing the where the response tends to be located for a particular value of the predictor and then smoothing that relationship out (hence the name).  We do want to point out that this is an exploratory device; we should be cautious about over-emphasizing relationships we observe from the spline.  Figure \@ref(fig:regsummaries-spline) illustrates a spline for the [Seismic Events Case Study](#CaseGreece).  The addition of the spline confirms what we had previously stated about the relationship appearing fairly linear (as the magnitude of the earthquake increases so does the bracketed duration at a location).  In addition to the spline, there is a confidence band around the line in order to convey the variability in the estimated smoothing spline.

```{r regsummaries-spline, echo=FALSE, fig.cap="Illustrating the use of a smoothing spline to explore the relationship between the bracketed duration and the magnitude of an earthquake for locations Greece."}
ggplot(data = greece.df,
       mapping = aes(x = Magnitude, y = BD02)) + 
  geom_point() +
  geom_smooth(method = "loess") +
  labs(x = "Moment Magnitude of the Earthquake",
       y = "Bracketed Duration (s) at Measurement Location") + 
  theme_bw(12)
```

As we have seen, supplementing graphical summaries with numerical summaries can help convey our message.  As an example, there is a positive linear relationship between the response and predictor in each of the cases illustrated in Figure \@ref(fig:regsummaries-correlation).  However, that relationship is much stronger or more apparent for Dataset A compared to Dataset B, for example.  It would be nice to have a numeric summary which captured this; such a metric is known as the __correlation coefficient__.

```{r regsummaries-correlation, echo=FALSE, fig.cap="Pairs of hypothetical variables which exhibit different correlations; that is, the relationship between each pair exhibit different strengths."}
set.seed(201709)
dat <- data_frame(
  dataset = rep(c("Dataset A", "Dataset B"), each = 100),
  sigma = (dataset=="Dataset A") + 5*(dataset=="Dataset B"),
  x = rep(rnorm(100), times = 2),
  y = 0 + 2*x + rnorm(200, mean = 0, sd = sigma)
)

ggplot(data = dat,
       mapping = aes(x = x, y = y)) +
  geom_point() +
  labs(x = "Predictor", y = "Response") +
  theme_bw(12) +
  theme(axis.ticks = element_blank(),
        axis.text = element_blank()) +
  facet_wrap(~ dataset)
```

```{definition, label=defn-correlation-coefficient, name="Correlation Coefficient"}
A numerical measure of the _strength_ and _direction_ of the _linear_ relationship between two quantitative variables.

The classical Pearson Correlation Coefficient $r$ is given by the following formula:
$$r = \frac{\sum_{i=1}^{n} \left(x_i - \bar{x}\right)\left(y_i - \bar{y}\right)}{\sqrt{\sum_{i=1}^n \left(x_i - \bar{x}\right)^2 \sum_{i=1}^n \left(y_i - \bar{y}\right)^2}}$$

where $\bar{x}$ and $\bar{y}$ represent the sample means of the predictor and response, respectively.
```

The correlation between the bracketed duration and the magnitude of an earthquake is `r round(cor(greece.df$BD02, greece.df$Magnitude), 2)`, indicating the two variables are positively linearly related, though perhaps the relationship is not strong.  

```{block2, type="rmdtip"}
Correlation coefficients measure both the strength and direction of linear relationships.  Here are a few of their key properties:

  - Takes a value between -1 and 1.
  - Negative values mean that the variables tend to move in opposite directions.
  - Positive values mean that the variables tend to move in the same direction.
  - Unitless and therefore unaffected by unit changes in the variables.

The biggest thing to remember is that a correlation coefficient measures the strength of a _linear_ relationship.  A correlation of 0 does not mean that two variables are unrelated.  It simply means they are not linearly related.
```


## Visualizing the Impact of a Third Variable on the Marginal Relationship
In the previous section, we stated that the bracketed duration tended to increase as the magnitude increased.  It is reasonable to ask the following question:

  > Is the relationship between the bracketed distance and the magnitude different depending on the soil condition of where the measurement is taken?
  
That is, we want to determine the impact that a third variable (soil condition) has on the relationship we have observed.  In order to add more depth to our graphical representations, we make use of various attributes of the graphic: color, shape, size, facets.  Figure \@ref(fig:regsummaries-color) uses color to distinguish between the three types of soil conditions for the measurement locations.  Notice the graphic allows us to both visualize the relationship for each soil condition but also facilitates our comparing these relationships.

```{r regsummaries-color, echo=FALSE, fig.cap="Relationship of the bracketed duration and the magnitude of an earthquake with various soil conditions."}
ggplot(data = greece.df,
       mapping = aes(x = Magnitude, y = BD02, 
                     colour = Soil_Condition)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Moment Magnitude of the Earthquake",
       y = "Bracketed Duration (s) at Measurement Location",
       colour = "Soil Conditions") +
  theme_bw(12) +
  scale_colour_brewer(palette = "Dark2", type = "qual") +
  theme(legend.position="bottom")
```

The figure illustrates that the relationship between magnitude and bracketed duration is similar for both locations which have soft or intermediate soil conditions.  However, for rocky conditions, the magnitude of the earthquake has a smaller impact on the resulting bracketed duration.  This suggests, possibly, that foundations on rocky soils are less subject to the effects of an earthquake.

While our focus has been on a different type of graphic, our emphasis remains consistent with the message of the overall text --- summaries need to be constructed to address the question of interest.
