# Using the Tools Together {#ANOVArecap}

This unit introduced a framework for determining if there is an association between a quantitative response and a categorical predictor.  We formed a standardized statistic for measuring the signal, and developed a model for the data-generating process which allowed us to model the null distribution of the standardized statistic.  In this chapter, we pull these tools together once more to answer a research question.

The primary question we have been addressing in this unit was whether the moral expectations of students were affected by the type of food to which they were exposed.  We saw that there was little evidence of a relationship between these two variables.  We now use the data from the [Organic Food Case Study](#CaseOrganic) to answer a related question:

  > Do the moral expectations of males and females differ?
  
  
## Framing the Question (Fundamental Idea I)
As stated, the above question is ill-posed.  We have not identified a variable or parameter of interest.  We refine this question to be

  > Does the average moral expectation score of males differ from that of females?
  
This question could also be stated as the following set of hypotheses:

  > Let $\mu_1$ and $\mu_2$ represent the average moral expectation score for males and females, respectively.  
  > $H_0: \mu_1 = \mu_2$  
  > $H_1: \mu_1 \neq \mu_2$
  

## Getting Good Data (Fundamental Idea II)
As we are working with previously collected data, our goal in this discussion is not how best to collect the data but making note of the limitations of the data as a result of how it was collected.  We previously described the [Organic Food Case Study](#CaseOrganic) as an example of a controlled experiment.  This was true...with regard to the primary question of interest (moral expectations and food exposure).  However, the subjects were _not_ randomly assigned to gender here; therefore, with regard to this question of interest, the data was an observational study.

It is common for young researchers to believe that if initially a controlled experiment was performed that the data always permits a causal interpretation.  However, we must always examine the data collection with respect to the question of interest.  Such "secondary analyses" (using data collected from a study to answer a question for which the data was not initially collected) are generally observational studies.  As a result, there may be other factors related to gender and moral expectations that drive any results we may see.


## Presenting the Data (Fundamental Idea III)
Our question here is examining the relationship between a quantitative response (moral expectation score) and a categorical predictor (gender).  Figure \@ref(fig:anovarecap-boxplot) compares the distribution of the moral expectation score for the two groups.  Note that `r sum(is.na(organic.df$gender))` students did not specify their gender; these subjects will be removed from the analysis.

```{r anovarecap-boxplot, echo=FALSE, fig.cap="Comparison of the moral expectations of males and females. The average value is added for each group.  Students who did not specify their gender were removed from the analysis."}
set.seed(201709)

ggplot(data = filter(organic.df, !is.na(gender)),
       mapping = aes(x = gender, y = moral_avg)) +
  geom_jitter(alpha = 0.25, width = 0.2, height = 0) +
  geom_boxplot(alpha = 0.25, size = 1.1) +
  stat_summary(fun.y = mean, geom = "point", size = 5, shape = 18) +
  labs(x = "", y = "Moral Expectation Score") +
  theme_bw(12)
```

Based on the above graphic, it appears the females tend to have higher moral expectations by about 1 point, compared to males.  We also observe that there are many more females in our sample compared to males, which is probably a result of the type of class and the demographic makeup of the university at which the study was conducted.


## Quantifying the Variability in the Estimate (Fundamental Idea IV)
In order to measure the size of the signal, we can compute the standardized statistic 
$$T = \frac{MSTrt}{MSE}$$

```{r anovarecap-models, echo=FALSE}
# Full Model
organic.gender.fullfit <- lm(moral_avg ~ gender, data = organic.df)

# Null Model
organic.gender.nullfit <- lm(moral_avg ~ 1, data = organic.df)

# ANOVA table
organic.gender.aov <- organic.gender.fullfit %>%
  anova() %>%
  tidy()
```

which is `r round(organic.gender.aov[1,5], 2)` for the sample we have observed.  Of course, if we were to collect a new sample, we would expect this value to change.  If we want to quantify the variability in this statistic, we need a model for its sampling distribution.  More, what we are really interested in is the sampling distribution of this statistic if the average moral expectation score were the same for the two genders; that is, we are interested in the null distribution of this standardized statistic.  With the null distribution, we could ascertain how unlikely our sample is to occur by chance alone (allowing us to quantify the strength of evidence in the data).

In order to model the null distribution, we consider the following model for the data-generating process:

\begin{equation}
  (\text{Moral Expectation Score})_i = \mu_1(\text{Male})_i + \mu_2 (\text{Female})_i + \epsilon_i
  (\#eq:anovarecap-model)
\end{equation}

where

$$
\begin{aligned}
  (\text{Male})_i &= \begin{cases}
    1 & \text{if i-th subject is male} \\
    0 & \text{if i-th subject is female} 
    \end{cases} \\
  (\text{Female})_i &= \begin{cases}
    1 & \text{if i-th subject is female} \\
    0 & \text{if i-th subject is male}
    \end{cases}
\end{aligned}
$$

and we impose the following conditions on the error term:

  1. The error in the moral expectation score for one individual is independent of the error in the moral expectation score for any other individual.
  2. The variance of the error in the moral expectation scores for males is the same as the variance of the error in moral expectation scores for females.
  3. The error in the moral expectation score for individuals follows a Normal Distribution.
  
Under these three assumptions, we are able to construct a model for the null distribution of the standardized statistic.  Figure \@ref(fig:anovarecap-classical-null-model) illustrates the null distribution assuming these conditions are satisfied.

```{r anovarecap-classical-null-model, echo=FALSE, fig.cap="Model for the standardized test statistic measuring the signal comparing the moral expectation scores for males and females in the Organic Food Case Study.  This model is constructed assuming the classical ANOVA conditions are satisfied."}
set.seed(201709)
null.dat <- data_frame(
  x = seq(0, 6.5, length.out = 1000),
  y = df(1000, 
         df1 = organic.gender.aov[1,"df"],
         df2 = organic.gender.aov[2, "df"])
)

ggplot(data = null.dat,
       mapping = aes(x = x, y = y)) +
  geom_blank() +
  stat_function(fun = df, args = list(df1 = organic.gender.aov[1, "df"],
                                      df2 = organic.gender.aov[2, "df"]),
                color = "blue", size = 1.1) +
  labs(x = "Standardized Test Statistic Assuming Null Hypothesis") +
  theme_bw(12) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank())
```

Before we can use this model to make any conclusions regarding our question of interest, we need to address the fact that we have assumed certain conditions are satisfied.  We need to assess whether the data is consistent with these assumptions.  This requires examining the residuals for the model.  

First, we discuss the assumption of independence.  Since the data was collected at a single point in time, known as a _cross-sectional study_, constructing a time-series plot of the residuals would not provide any information regarding this assumption.  Instead, we rely on the context of the problem to make some statements regarding whether the data is consistent with this condition (whether making this assumption is reasonable).  It is reasonable that the errors are independent.  One case in which this might be violated is if students discussed their answers to the questions as they filled out the survey; then, it is plausible that one student influenced another's responses.  As this is unlikely given the description of the data collection, we feel it is reasonable to assume independence. 

Again, note that there is a condition of independence; we are simply saying whether we are willing to assume the condition is satisfied.  There is no way to ensure the condition holds.

In order to assess the condition of constant variance, let us look back at the box plots given in Figure \@ref(fig:anovarecap-boxplot).  As the spread of the moral expectation score for each of the two genders is roughly the same, it is reasonable to assume the variability of the errors in each group is the same.

Finally, to assess the condition that the distribution of the errors is Normal, we consider a probability plot of the residuals (Figure \@ref(fig:anovarecap-resids-probplot)).  Given that the residuals tend to display a linear relationship, it is reasonable that the residuals represent a sample from a Normal Distribution.  That is, it is reasonable that the errors follow a Normal Distribution.

```{r anovarecap-resids-probplot, echo=FALSE, fig.cap="Probability plot assessing the assumption that the errors for our model comparing the moral expectation score across gender follow a Normal Distribution."}
# Obtain Residuals
organic.gender.aug <- suppressWarnings(augment(organic.gender.fullfit))

ggplot(data = organic.gender.aug,
       mapping = aes(sample = .resid)) +
  stat_qq() +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_bw(12)
```

Given that we are comfortable assuming the conditions on the error term are reasonable, we can make use of the analytical model for the null distribution in Figure \@ref(fig:anovarecap-classical-null-model).


## Quantifying the Evidence (Fundamental Idea V)
Now that we have a model for the null distribution, we can determine how extreme our particular sample was by comparing the standardized statistic for our sample with this null distribution.  We can measure this through computation of a p-value, the probability that we would observe a standardized statistic of this magnitude or higher by chance alone if there were no difference in the mean moral expecation scores of males and females.  This is summarized in Table \@ref(tab:anovarecap-anova-table) below.

```{r anovarecap-anova-table, echo=FALSE}
organic.gender.aov %>%
  rename(Source = term,
         DF = df,
         SS = sumsq,
         MS = meansq,
         F = statistic,
         `P-value` = p.value) %>%
  mutate(Source = recode(Source,
                         "gender" = "Gender")) %>%
  add_row(Source = "Total",
          DF = sum(.$DF),
          SS = sum(.$SS)) %>%
  knitr::kable(digits = 3, caption = "ANOVA table summarizing the comparison of the moral expectation score across gender within the Organic Food Case Study.")
```

From the results, we can conclude that there is evidence (p = `r round(organic.gender.aov[1,6], 3)`) of a relationship between the moral expectations of a student and their gender.  Looking back at Figure \@ref(fig:anovarecap-boxplot), females tend to have higher moral expectations on average.


## Conclusion
Throughout this unit, we have examined a framework for examining the association between a quantitative response and a categorical predictor.  This reinforces a couple of big ideas we have seen throughout this text:

  - The key to measuring a signal is to partition the variability in the response.
  - A standardized statistic is a numeric measure of the signal strength in the sample.
  - Modeling the data-generating process provides us a way of modeling the sampling distribution of the parameter estimates and the null distribution of a standardized statistic.
  - Conditions are often placed on the noise portion of the model for the data-generating process; before assuming these conditions are met, we should graphically assess whether the data is consistent with these conditions.

