# Assessing Modeling Assumptions {#Blockassessment}

In this unit we have discussed a model relating a quantitative response to a categorical predictor in the presence of blocks which induce correlation among the responses.  For the [Frozen Yogurt Case Study](#CaseYogurt), our model had the form

$$
\begin{aligned}
  (\text{Taste Rating})_i &= \mu_1 (\text{East Side})_i + \mu_2 (\text{South Side})_i + \mu_3 (\text{Name Brand})_i \\
    &\qquad + \beta_2 (\text{Subject 2})_i + \beta_3 (\text{Subject 3})_i + \beta_4 (\text{Subject 4})_i \\
    &\qquad + \beta_5 (\text{Subject 5})_i + \beta_6 (\text{Subject 6})_i + \beta_7 (\text{Subject 7})_i \\
    &\qquad + \beta_8 (\text{Subject 8})_i + \beta_9 (\text{Subject 9})_i + \epsilon_i
\end{aligned}
$$

where

$$
\begin{aligned}
  (\text{East Side})_i &= \begin{cases}
    1 & \text{if i-th rating associated with east side yogurt vendor} \\
    0 & \text{if i-th rating not associated with east side yogurt vendor}
    \end{cases} \\
  (\text{South Side})_i &= \begin{cases}
    1 & \text{if i-th rating associated with south side yogurt vendor} \\
    0 & \text{if i-th rating not associated with south side yogurt vendor}
    \end{cases} \\
  (\text{Name Brand})_i &= \begin{cases}
    1 & \text{if i-th rating associated with name brand yogurt vendor} \\
    0 & \text{if i-th rating not associated with name brand yogurt vendor}
    \end{cases} \\
  (\text{Subject j})_i &= \begin{cases}
  1 & \text{i-th observation taken from Subject j} \\
  0 & \text{i-th observation not taken from subject j}
  \end{cases}
\end{aligned}
$$

Further, we considered three conditions on the distribution of the error term:

  1.  The error in the taste ratings within one individual is independent of the error in the taste ratings within any other individual.
  2.  The error for the taste rating is identically distributed.
  3.  The error in the taste rating follows a Normal distribution.
  
In addition, we considered four conditions on the distribution of the random effects:

  1.  Each subject's effect is identical for all vendors.
  2.  The effect of one subject is independent of the effect of any other subject.
  3.  The effect of each subject is independent of the error in the taste ratings.
  4.  The effect of each subject follows a Normal distribution.
  
We make a distinction in these conditions because the latter four conditions are not easily assessed.  These are generally stated and assumed.  However, we can assess the conditions on the error in a similar way to previous units.  Just as in Chapters \@ref(Regassessment) and \@ref(ANOVAassessment), we use residuals to assess these conditions qualitatively.


## Assessing Independence
  > The error in the taste ratings within one individual is independent of the error in the taste ratings within any other individual.
  
When a study design incorporates a block, it is rare that the researchers report the order in which the data was collected.  This tends to be because the data is collected in the most convenient manner, as opposed to randomizing the order of the data collection.  In such cases, we must really rely on the context to determine if it is reasonable to assume this condition.  When the order is known, we can use a time-series plot of the residuals. As in the past, this particular plot only assesses one particular deviation from this condition, but it is the most common.

If two errors were unrelated, then the value of one residual should tell us nothing about the value of the next residual.  Therefore, a plot of the residuals over time should look like noise (since residuals are supposed to be estimates of noise).  If there are any trends, then it suggests the data is not consistent with independence.  Of course, to construct such a plot, we need to know the order in which the data was collected.

For the [Frozen Yogurt Case Study](#CaseOrganic), participants were assessed simultaneously within the class.  Therefore, there is no ordering in time to be concerned about.  As such, a time-series plot of the residuals would not be useful here.  Considering the context, the students were the ones who had designed the study (it was carried out by their instructor); therefore, they were concerned about the quality of the data collection.  The students did their best to not influence any other's ratings.  It is reasonable to assume the data is consistent with this condition.


## Assessing Identical Distribution
We want to assess the variability in the residuals in order to determine if the data is consistent with the errors having an identical distribution.  Unlike Chapter \@ref(ANOVAassessment), we cannot simply make a boxplot of the errors across the three groups.  Once there is more than a single predictor in the model, we resort back to the general strategy discussed in the regrssion unit and construct a plot of the residuals against the predicted values.  Figure \@ref(fig:blockassessment-variance-yogurt) shows the residuals for each individual across the predicted taste rating for that individual observation.  Notice that as the predicted taste ratings increase, the variability in the residuals tends to decrease.  This would suggest the data is not consistent with this condition.  If it were, we would not expect to see this fan shape.

```{r blockassessment-variance-yogurt, echo=FALSE, fig.cap="Comparison of the residuals from comparing the taste ratings across yogurt vendors accounting for subject-variability."}
yogurt.augmented = augment(fit.yogurt)

ggplot(data = yogurt.augmented,
       mapping = aes(y = .resid,
                     x = .fitted)) +
  geom_point() +
  labs(y = "Residuals",
       x = "Predicted Taste Ratings") +
  theme_bw(12)
```


This plot does suggest the residuals tend to be centered around 0.  You may think "hold on, I don't remember discussing the mean-0 condition."  We did this indirectly.  When we assume that the random effect for each subject is the same across the vendors, this is essentially saying that the effect of the vendor is not dependent upon the subject we are working with (no interaction, see Definition \@ref(def:defn-interaction)).  Since our model does not allow for the effect of the vendor to depend upon the subject, we are making an assumption about the form of the model, which is exactly what the mean-0 condition in regression does.  While we have used residuals to assess this condition, we can also assess this based on context in the case of blocks.  It is reasonable to assume that the average taste ratings for a particular vendor are similar across subjects.


## Assessing Normality
Just as before, we emphasize the probability plot as a method for assessing whether data is consistent with being drawn from a Normal distribution.

Figure \@ref(fig:blockassessment-normal-yogurt) shows the probability plot for the residuals from the [Frozen Yogurt Case Study](#CaseYogurt).

```{r blockassessment-normal-yogurt, echo=FALSE, fig.cap="Probability plot of the residuals for the Frozen Yogurt Case Study.  If the errors follow a Normal distribution, we would expect the residuals to fall along a straight line."}
ggplot(data = yogurt.augmented,
       mapping = aes(sample = .resid)) +
  stat_qq() +
  labs(x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_bw(12)
```

Overall, the points do tend to follow a straight line.  It seems reasonable the data is consistent with the errors following a Normal distribution.


## General Tips for Assessing Assumptions
As in Chapter \@ref(Regassessment), we want to point out three things that should be kept in mind when assessing these conditions:

  1. We can never prove a condition is satisfied; we can only determine whether the data is consistent with a condition or whether it is not consistent with a condition.
  2. The analysis has specific conditions placed on the error term; we choose whether to assume such a condition is reasonable by examining residuals.
  3. A sample should always be consistent with the conditions you are relying on to interpret a p-value or confidence interval.
  
