# Case Study: Paper Strength {#CasePaper}

While electronic records have become the predominant means of storing information, we do not live in a paperless society.  Paper products are still used in a variety of applications ranging from printing reports and photography to packaging and bathroom tissue.  In manufacturing paper for a particular application, the strength of the resulting paper product is a key characteristic. 

There are several metrics for the strength of paper.  A conventional metric for assessing the inherent (not dependent upon the physical characteristics, such as the weight of the paper, which might have an effect) strength of paper is the _breaking length_.  This is the length of a paper strip, if suspended vertically from one end, that would break under its own weight.  Typically reported in kilometers, the breaking length is computed from other common measurements.  For more information on paper strength measurements and standards, see the following website: <http://www.paperonweb.com>

A study was conducted at the University of Toronto to investigate the relationship between pulp fiber properties and the resulting paper properties [@Lee1992].  The breaking length was obtained for each of the 62 paper specimens, the first 10 measurements of which are shown in Table \@ref(tab:casepaper-table).  The complete data is available online at the following website: <https://vincentarelbundock.github.io/Rdatasets/doc/robustbase/pulpfiber.html>

While there are several question one might ask with the available data, we must first begin discuss how we might characterize the breaking length of these paper specimens.  We consider various techniques for summarizing a variable.

```{r casepaper-data, echo=FALSE}
# Obtain Data
#  Available online at http://users.ugent.be/~svaelst/data/pulpfiber.txt.
paper.df <- read_table2(
  "http://users.ugent.be/~svaelst/data/pulpfiber.txt",
  col_names = c("Specimen",
                "Arithmetic_Fiber_Length",
                "Long_Fiber_Fraction",
                "Fine_Fiber_Fraction",
                "Zero_Span_Tensile",
                "Breaking_Length",
                "Elastic_Modulus",
                "Stress_at_Failure",
                "Burst_Strength"),
  col_types = "innnnnnnn",
  skip = 1
)
```

```{r casepaper-table, echo=FALSE}
# Print Table
#  Reduce to only show variable of interest.
paper.df %>%
  select(Specimen,
         `Breaking Length` = Breaking_Length) %>%
  slice(1:10) %>%
  knitr::kable(caption = "Breaking length (km) for first 10 specimens in the Paper Strength study.")
```
