# (PART) Unit III: Modeling the Average Response as a Function of Several Predictors {-}

# Case Study: Seismic Activity in Greece {#CaseGreece}

At the intersection of the African plate, the Eurasia plate, and the smaller Aegean plate, Greece is one of the most earthquake-prone regions in the world.  Between July 2016 and July 2017, Greece experienced 179 earthquakes; by contrast, the state of Texas experienced 28 over the same span of time.  In a region with such seismic activity, careful consideration must be given to municipal construction.  Further, understanding how the motion experienced in a location is related to the soil properties in the area or the magnitude and distance of an earthquake is important.

An article in the _Journal of Earthquake Engineering_ [@Koutrakis2002] examined seismic events in Greece occurring between 1978 and 1997.  Of interest for construction is characterizing the "strong ground motion," when the earth shakes with enough force to cause damage to infrastructure, with respect to the properties of a location.  The study recorded several measurements for a collection of 121 seismic events[^caveat].  The primary variable of interest is the _uniform duration_, the amount of time, in seconds, the ground acceleration (which occurs when the earth is shaking) exceeds a specified value.  For this study, the specified threshold was twice the acceleration due to gravity.  In addition, the following measurements were available for each observation:

- magnitude of the nearby earthquake on the Richter scale (larger values indicate more severe earthquakes)
- distance in kilometers of the measurement from the epicenter of the earthquake
- indicator (Yes/No) of whether the local soil consists of soft alluvial deposits (alluvial deposits consist fine particles of silt and clay as well as larger particles of gravel and sand).
- indicator (Yes/No) of whether the local soil consists of pre-Quaternary rocks (rocks older than 2.58 million years) 

The first 5 observations in the dataset are shown in Table \@ref(tab:casegreece-table).  While there are several question one might ask with the available data, we first discuss how we might characterize the uniform duration of these events; that is the emphasis of Part I of the text.

```{r casegreece-data, echo=FALSE}
# Obtain Data
#  Available originally from Navidi, Ch 8, Supplementary Exercise 22.
greece.df <- 
  read_csv("./data/case-greece.csv",
           col_types = "nniii") %>%
  mutate(Soil_Soft_Alluvial_Deposits = factor(Soil_Soft_Alluvial_Deposits,
                                              levels = c(0, 1),
                                              labels = c("No", "Yes"),
                                              ordered = TRUE),
         Soil_Quaternary_Rock = factor(Soil_Quaternary_Rock,
                                       levels = c(0, 1),
                                       labels = c("No", "Yes"),
                                       ordered = TRUE))
```

```{r casegreece-table, echo=FALSE}
greece.df %>%
  rename(`Uniform Duration (s)` = Uniform_Duration,
         `Distance from Epicenter (km)` = Distance_from_Epicenter,
         `Soft Alluvial Deposits Present` = Soil_Soft_Alluvial_Deposits,
         `Quaternary Rock Present` = Soil_Quaternary_Rock) %>%
  slice(1:5) %>%
  knitr::kable(caption = "Data for first 5 observations from study characterizing seismic activity in Greece.")
```

[^caveat]: The original article presented repeated measurements at some locations; these have been excluded from the dataset presented here to simplify any analyses.  Repeated measurements are discussed briefly later in the text; for a more thorough treatment of the subject, we recommend a course in Designed Experiments or Biostatistics.  The dataset presented here corresponds to that presented in Navidi's "Statistics for Engineers and Scientists" (Chapter 8, Supplementary Exercise 22).
