# Asking the Right Questions {#Questions}

The discipline of statistics is about turning data into information in order to address some question.  While there may be no such thing as a stupid question, there are most certainly ill-posed questions --- those which cannot be answered as stated.  Consider the [Seismic Activity Case Study](#CaseGreece).  It might seem natural to ask "does the uniform duration exceed 5 s for seismic events in Greece?"  However, we quickly see that this is an ill-posed question.  The first observation in the sample has a uniform duration of 8.82 s, while the second observation has a uniform duration of only 4.08 s.  The answer to our question, therefore, seems to be "sometimes."  In fact, looking at a plot of the data (Figure \@ref(fig:questions-greece-dotplot)) shows that the uniform duration for those observations in our sample can be quite different.

```{r questions-greece-data, echo=FALSE, ref.label="casegreece-data"}
```

```{r questions-greece-dotplot, echo=FALSE, fig.cap="Uniform duration (s) for 121 seismic events in Greece."}
ggplot(data = greece.df,
       mapping = aes(x = Uniform_Duration)) +
  geom_dotplot(binwidth = 0.75, method="histodot") +
  labs(x = "Uniform Duration of Event (seconds)") +
  theme_bw(12) +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())
```

So, __variability__ makes some questions ill-posed.  It is variability that creates a need for statistics; in fact, you could think of statistics as the study and characterization of variability.  We must therefore learn to ask the _right_ questions --- those which can be answered in the presence of variability.

```{definition, label=defn-variability, name="Variability"}
The notion that measuremenets differ from one observation to another.
```

```{block2, type="rmdkeyidea"}
The presence of variability makes some questions ill-posed; statistics concerns itself with how to address questions in the presence of variability.
```


## Characterizing a Variable
Since we want to say something about the population, any question we ask should then be centered on this larger group.  The first step to constructing a well-posed question is then to identify the population of interest for our study.  For the [Seismic Activity Case Study](#CaseGreece), it is unlikely that we are only interested in these 121 observed events between 1978 and 1997.  In reality, the 121 observations in our dataset form the sample, a subset from all seismic events that occur in Greece.  That is, our population of interest is comprised of all seismic events in Greece (both past and future).

```{block2, type="rmdtip"}
When identifying the population of interest, be specific!  Are you really interested in _all_ trees, for example?  Or, are you interested in Maple trees within the city limits of Terre Haute, Indiana?
```

Since we can expect that the uniform duration (the variable of interest for this study) varies from one seismic event to the next, we cannot ask a question about the _value_ of the uniform duration itself (since the value differs for each subject).  Instead, we want to characterize the __distribution__ of the uniform duration.

```{definition, label=defn-distribution, name="Distribution"}
The pattern of variability corresponding to a set of values.
```

There are many ways to characterize a distribution, many of which are discussed in the chapter [Summarizing the Evidence](#Summaries).  The most common such characterization is the mean, or average.  Such numeric quantities which summarize the distribution of the variable within the population are known as __parameters__.

```{definition, label=defn-parameter, name="Parameter"}
Numeric quantity which summarizes the distribution of a variable within the _population_ of interest.
```

While the _value_ of a variable may vary across the population, the _parameter_ is a single fixed constant which summarizes the variable for the population.  Therefore, well-posed questions can be constructed if we limit ourselves to questions about the parameter.  The second step in constructing well-posed questions is then to identify the parameter of interest.  

The questions we ask then generally fall into one of two categories:

  - Estimation: what is the _average_ uniform duration for seismic events in Greece?
  - Model Consistency: is it reasonable the _average_ uniform duration is similar for sesimic events regardless of whether the soil consists of alluvial deposits, or is there evidence that the _average_ uniform duration differs between these two types of soil?

Now, since we do not get to observe the population (we only see the sample), we cannot observe the value of the parameter.  That is, we will never know the average uniform duration for all seismic events in all locations in Greece.  However, we can use the sample to inform us of what the data suggests.

```{definition, label=defn-estimation, name="Estimation"}
Using the sample to approximate the value of a parameter from the underlying population.
```

```{definition, label=defn-hypothesis-testing, name="Hypothesis Testing"}
Using a sample to determine if the data is consistent with a working theory or if there is evidence to suggest the data is not consistent with the theory.
```

```{block2, type="rmdkeyidea"}
Parameters are unknown values and can, in general, never be known.  Parameters are generally denoted by Greek letters in statistical formulas.
```

It turns out, the vast majority of research questions can be framed in terms of a parameter.  In fact, this is the first of what we consider the _Five Fundamental Ideas of Inference_.

```{block2, type="rmdfivefund"}
__Fundamental Idea I__: A research question can often be framed in terms of a parameter which characterizes the population.  Framing the question should then guide our analysis.
```


## Framing the Question
In engineering and scientific applications, many questions fall under the second category of model consistency.  Examining such questions is known as __hypothesis testing__.  Specifically, data is collected to help the researcher choose between two competing theories for the parameter of interest.  In this section, we consider the terminology surrounding specifying such questions.

For the [Seismic Activity Case Study](#CaseGreece) suppose we are interested in addressing the following question:

  > Is there evidence that the average uniform duration for seismic events in Greece exceeds 5 seconds?
 
The question itself is about the population (seismic events in Greece) and is centered on a parameter (the average uniform duration).  That is, this is a well-posed question that can be answered with appropriate data.  The overall process for addressing these types of questions is similar to conducting a trial in a court of law.  In the United States, a trial has the following essential steps:

1. Assume the defendant is innocent.
2. Evidence to establish guilt (to the contrary of innocence) is presented by the prosecution.
3. The jury considers the weight of the evidence.
4. If the evidence is "beyond a reasonable doubt," the jury declares the defendant guilty; otherwise, the jury declares the subject not guilty.

The process of conducting a hypothesis test has similar essential steps:

1. Assume the innocent of what we want the data to show (develop a working theory).
2. Gather data and compare it to the proposed model.
3. Quantify the likelihood of our data under the proposed model.
4. If the likelihood is small, conclude the data is not consistent with the working model (there is evidence for what we want to show); otherwise, conclude the data is consistent with the working model (there is no evidence for what we want to show).

Notice that a trial centers not on proving guilt but on disproving innocence; similarly, in statistics, we are able to establish evidence _against_ a specified theory.  This overall process has several other subtle points, and it is okay if the process is not yet clear.  Future chapters will discuss methods for actually implementing this process and discuss each step in greater detail.  Here, we focus solely on that first step --- developing a working theory that we want to disprove.

Consider the above question for the [Seismic Activity Case Study](#CaseGreece).  We want to show that the average uniform duration exceeds 5 s.  Therefore, we would like to _disprove_ (or provide evidence _against_) the statement that the average uniform duration for seismic events in Greece is no more than 5 s.  This is known as the __null hypothesis__; the opposite of this statement, called the __alternative hypothesis__, captures what we would like to establish.

```{definition, label=defn-null-hypothesis, name="Null Hypothesis"}
The statement (or theory) that we would like to _disprove_.  This is denoted $H_0$, read "H-naught" or "H-zero".
```

```{definition, label=defn-alternative-hypothesis, name="Alternative Hypothesis"}
The statement (or theory) capturing what we would like to provide evidence _for_; this is the opposite of the null hypothesis.  This is denoted $H_1$ or $H_a$, read "H-one" and "H-A" respectively.
```

For the [Seismic Activity Case Study](#CaseGreece), we write:

  > $H_0:$ The average uniform duration of seismic events in Greece is no more than 5 s.  
  > $H_1:$ The average uniform duration of seismic events in Greece exceeds 5 s.

Each hypothesis is a well-posed statement (about a parameter characterizing the entire population).  The two statements are exactly opposite of one another; only one can be a true statement.   We can now collect data to determine if it is consistent with the null hypothesis (a statement similar to "not guilty") or if the data provides evidence against the null hypothesis and in favor of the alternative (a statement similar to "guilty").

Often these statements are written in a bit more of a mathematical structure in which a Greek letter is used to represent the parameter of interest.  For example, we might write

  > Let $\theta$ be the average uniform duration, in seconds, for seismic events in Greece.  
  > $H_0: \theta \leq 5$  
  > $H_1: \theta > 5$

In the above statements, $\theta$ represents the parameter of interest; the value 5 is known as the __null value__.

```{definition, label=defn-null-value, name="Null Value"}
The value associated with the equality component of the null hypothesis; it forms the threshold or boundary between the two hypothesis.  Note: not all questions of interest require a null value be specified.
```

```{block2, type="rmdkeyidea"}
Hypothesis testing is a form of statistical inference in which we quantify the evidence _against_ a working theory (captured by the null hypothesis).  We essentially argue that the data supports the alternative if it is not consistent with the working theory.
```

```{block2, type="rmdtip"}
__Process for Framing a Question__ In order to frame a research question, consider the following steps:

1. Identify the population of interest.
2. Identify the parameter(s) of interest.
3. Determine if you are interested in estimating the parameter(s) or quantifying the evidence against some working theory.
4. If you are interested in testing a working theory, make the null hypothesis the working theory and the alternative the exact opposite statement (what you want to provide evidence for).

```
